---
title: "Policy Wheel Tutorial"
author: "Joshua Eagan, Max Griswold, Beth Ann Griffin, Seema (Choksy) Pessar, Rosalie Pacula, Bradley Stein"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

# Introduction

This tutorial demonstrates the creation of OPTIC style policy wheel data visualizations. These graphs are meant to illustrate the timeline over which state level policies are implemented concurrently. For some examples of these data visualizations, skip to the middle of this tutorial.

-------------------------------------------

# Setup

To create one of these graphs, you first need to load in a few R packages: 

* `circlize` (Gu, Z., 2014)
* `data.table` (Srinivasan, 2023)
* `plyr` (Grolemund, 2011)
* `lubridate` (Grolemund, 2011)

You also need to load in the appropriate functions from this project's working directory- we did this below using `source`.


```{r message=FALSE}

# loading packages

suppressPackageStartupMessages(library(circlize))
library(data.table)
library(plyr)
library(lubridate)

# loading functions
for(file in list.files("R")){
  cat(paste0(file, "\n") )
  source(paste("R", file, sep="/"))
}

```

The R function we use to plot policy wheels is `plot_policy_wheels`. 

-------------------------------------------

# Plotting Policy wheels

Next, you need to load in your data. In the code below, we load in a wide version of the data, meaning that there is one row for each state and a column for each policy, containing the date each policy was implemented for each state. Next, we rename columns in the wide data so variables are named with the text we want displayed in the graphs.

```{r}

# reading in a data frame where rows are states, columns are policies, and values are enactment dates

# Dates can come in the forms: "1/15/2015", "2015-01-15", "2015", or "January 15, 2015"
df_wide <- read.csv('Data/processed/example_data_wide.csv')
names(df_wide) = gsub("\\.", " ", names(df_wide))

# renaming variables
names(df_wide)[names(df_wide) == "nal_date_any_nal"] = "Any Naloxone Access Law (NAL)"
names(df_wide)[names(df_wide) == "nal_date_nal_protocol_standing"] = "NAL Standing Order or Protocol"
names(df_wide)[names(df_wide) == "nal_date_nal_Rx_prescriptive_auth"] = "NAL Prescriptive Authority"

# All NAL laws and GSL
names(df_wide)[names(df_wide) == "gsl_date_anygsl"] = "Any Good Samaritan Law (GSL)"
names(df_wide)[names(df_wide) == "gsl_date_gsl_arrest"] = "GSL Arrest"

# NAL laws and Medical/rec cannabis access 
names(df_wide)[names(df_wide) == "mm_date_effMML"] = "Medical Marijuana Law"
names(df_wide)[names(df_wide) == "mm_date_active_medlegdisp"] = "Medical MJ Dispensaries Open"
names(df_wide)[names(df_wide) == "mm_date_effREC"] = "Recreational Marijuana Law"
names(df_wide)[names(df_wide) == "mm_date_active_dispREC"] = "Recreational MJ Store Open"

# Co-prescribing
names(df_wide)[names(df_wide) == "copnal_date_all_prescribe"] = "Naloxone Co-prescribing\nMandating Prescription"

# looking at the data
head(df_wide)

```


Once the data are clean, you can run `plot_policy_wheels` to make your graph.


```{r message=FALSE}

# generating policy wheels
plot_policy_wheels(data = df_wide,
                   
                   # Ordering policies by name:
                  policies = c("Any Naloxone Access Law (NAL)", "NAL Standing Order or Protocol", "NAL Prescriptive Authority", "Any Good Samaritan Law (GSL)", "GSL Arrest"),
                  
                  # name of the state variable
                  state_var = "state",
                  
                  # Restrict to relevant policy intervals, for locations that implemented the policy
                  policy_intervals = c(2010, 2015, 2020),
                  plot_colors = c("#1f77b4", "#ff7f0e", "#FFFF00", "#dab8e5", "#9467bd"),
                  legend_args = list(x = "center", xjust = 0.5, y.intersp = 1.3, x.intersp = 1.3, cex = 2.5, pt.cex = 2.7, bty = "n", ncol = 2),
                  
                  panel_width = 4,
                  panel_height = 5,
                  
                  # where should the graph be saved?
                  out_file = "www/policy_wheel_1.svg")

# displaying the new graph
knitr::include_graphics("www/policy_wheel_1.svg")

```

Referencing the graph, you can see that the ordering of the `policies` argument corresponds to the ordering of `plot_colors`. The policy enactment dates in `df` were converted to years and split across 3 separate wheels according to `policy_intervals`. There is an interesting amount of variation in the "in-between" years (2011-2014 and 2016-2019)- so increasing the amount of years displayed might better illustrate these trends. You can see an example of this change below.

```{r message=FALSE}

# generating policy wheels
plot_policy_wheels(data = df_wide,
                   
                   # Ordering policies by name:
                  policies = c("Any Naloxone Access Law (NAL)", "NAL Standing Order or Protocol", "NAL Prescriptive Authority", "Any Good Samaritan Law (GSL)", "GSL Arrest"),
                  
                  # name of the state variable
                  state_var = "state",
                  
                  # Restrict to relevant policy intervals, for locations that implemented the policy
                  policy_intervals = c(2010, 2013, 2016, 2019, 2021),
                  plot_colors = c("#1f77b4", "#ff7f0e", "#FFFF00", "#dab8e5", "#9467bd"),
                  legend_args = list(x = "center", xjust = 0.5, y.intersp = 1.3, x.intersp = 1.3, cex = 2.5, pt.cex = 2.7, bty = "n", ncol = 2),
                  
                  panel_width = 4,
                  panel_height = 5,
                  
                  nrows = 2,
                  ncols = 3,
                  
                  # where should the graph be saved?
                  out_file = "www/policy_wheel_1_revised.svg")

# displaying the new graph
knitr::include_graphics("www/policy_wheel_1_revised.svg")

```

If you are making your own plots, you may can also edit the following settings:


* `plot_width` and `plot_height`: these edit the dimensions of the plot output and will need to be adjusted according to your needs.
* `legend_args`: this is a list whose elements will be passed to `graphics::legend()`. Within the function, two arguments are already configured for you, `legend` and `col`, but it's up to you to configure the remaining. The defaults are probably fine here, but if you want to change anything, refer to `?graphics::legend`.

See another example of a policy wheel below:

```{R message=FALSE}

# generating policy wheels
plot_policy_wheels(data = df_wide,
                   
                   # Ordering policies by name:
                  policies = c("Any Naloxone Access Law (NAL)", "NAL Standing Order or Protocol", "NAL Prescriptive Authority",
                    "Medical Marijuana Law", "Medical MJ Dispensaries Open", "Recreational Marijuana Law", "Recreational MJ Store Open"),
                  
                  # name of the state variable
                  state_var = "state",
                  
                  # Restrict to relevant policy intervals, for locations that implemented the policy
                  policy_intervals = c(2012, 2015, 2018, 2021),
                  plot_colors = c("#1f77b4", "#ff7f0e", "#FFFF00", 
                                  "#f7a8a8", "#d62728", "#90EE90","#006400"),
                  legend_args = list(x = "center", xjust = 0.5, y.intersp = 1.3, x.intersp = 1.3, cex = 2.5, pt.cex = 2.7, bty = "n", ncol = 2),
                  
                  plot_width = 22, 
                  plot_height = 16,
                  
                  nrows = 2,
                  ncols = 2,
                  panel_width = 5,
                  panel_height = 6,
                  
                  # where should the graph be saved?
                  out_file = "www/policy_wheel_2.svg")

# displaying the new graph
knitr::include_graphics("www/policy_wheel_2.svg")


```

In this example, we include data for 7 policies at 4 time points in 3 year intervals. Also, note that the `nrows` and `ncols` arguments were used to control the layout of the wheels. The result is a bit crowded, and we would recommend including fewer than 7, or even 6 policies in one of these plots. We can infer that most states effected a Naloxone Access Law (NAL) between 2012 and 2015 (blue). Many of these laws allowed the distribution of Naloxone through a standing or protocol order (orange), but most states did not pass laws allowing pharmacists prescriptive authority (yellow) until later. We can also see that the states in the West and North-East tended to pass these policies across the board earlier than states in the Midwest and South. Also notice that medicinal (red) and recreational (green) marijuana laws were passed around the same time (if not slightly after) NAL in many states.

Lastly, create a plot that shows the association between the passage of NAL laws and Naloxone Co-prescribing Laws that mandate prescribing and affect all patients.

```{R message=FALSE}

# generating policy wheels
plot_policy_wheels(data = df_wide,
                   
                   # Ordering policies by name:
                  policies = c("Any Naloxone Access Law (NAL)", "NAL Standing Order or Protocol", "NAL Prescriptive Authority",
                    "Naloxone Co-prescribing\nMandating Prescription"),
                  
                  # name of the state variable
                  state_var = "state",
                  
                  # Restrict to relevant policy intervals, for locations that implemented the policy
                  policy_intervals = c(2012, 2014, 2016, 2018, 2020, 2022),
                  plot_colors = c("#1f77b4", "#ff7f0e", "#FFFF00", 
                                  "#00b3b3"),
                  legend_args = list(x = "center", xjust = 0.5, y.intersp = 1.3, x.intersp = 1.3, cex = 2.5, pt.cex = 2.7, bty = "n", ncol = 2),
                  
                  plot_width = 22, 
                  plot_height = 16,
                  
                  panel_width = 5,
                  panel_height = 6,
                  
                  # where should the graph be saved?
                  out_file = "www/policy_wheel_3.svg")

# displaying the new graph
knitr::include_graphics("www/policy_wheel_3.svg")


```

-------------------------------------------

# Sources:

* Gu, Z. circlize implements and enhances circular visualization in R. Bioinformatics 2014.
* Garrett Grolemund, Hadley Wickham (2011). Dates and Times Made Easy with lubridate. Journal of Statistical Software, 40(3), 1-25. URL https://www.jstatsoft.org/v40/i03/
* Dowle M, Srinivasan A (2023). _data.table: Extension of `data.frame`_. R package version 1.14.8, <https://CRAN.R-project.org/package=data.table>
* Hadley Wickham (2011). The Split-Apply-Combine Strategy for Data Analysis. Journal of Statistical Software, 40(1), 1-29. URL https://www.jstatsoft.org/v40/i01/.
